% !Mode:: "TeX:UTF-8"
\hitsetup{
  statesecrets={公开},
  ctitlecover={非线性系统预设阻尼比滑模反步控制方法研究},
  ctitle={非线性系统预设阻尼比滑模反步控制方法研究},
  cxueke={工学},
  csubject={控制科学与工程},
  caffil={航天学院},
  cauthor={佘宇琪},
  csupervisor={郑晓龙},
  cstudentid={23S004010},
  cstudenttype={学术学位论文},
  cdate={2025年6月},
  etitle={Backstepping Control With Predefined Damping Ratios: A Second-Order Lyapunov Method},
  exueke={Engineering},
  esubject={Control Science and Engineering},
  eaffil={\emultiline[t]{School of Astronautics\\Harbin Institute of Technology}},
  eauthor={Yuqi She},
  esupervisor={Professor Xiaolong Zheng},
  edate={June, 2025},
  estudenttype={Master Thesis},
  ckeywords={预设阻尼比, 滑模反步控制, 神经网络, 指令滤波抗扰, 抗饱和},
  ekeywords={Predefined damping ratio, Sliding-mode backstepping, Second-order Lyapunov method, Neural-network compensation, Second-order command-filtered anti-disturbance, Anti-saturation control},
}

\begin{cabstract}
针对不确定非线性系统中瞬态性能难以直接设计的问题，本文以严格反馈系统为研究对象，在滑模反步框架下研究预设阻尼比控制方法。论文以二阶类李雅普诺夫方法为主线，将输出误差的收敛过程约束为满足二阶微分不等式，使闭环主导动态能够由控制器参数直接塑形，从而把传统反步控制中难以显式整定的瞬态指标转化为可解释、可调节的阻尼比参数设计问题。

首先，针对一类严格反馈不确定非线性系统，建立二阶类李雅普诺夫稳定性判据及其参数选取规则给出预设阻尼比滑模反步控制方法。通过设计积分滑模面、虚拟控制器和实际控制律，构造包含主导阻尼参数的闭环误差系统，输出响应可近似保持预设阻尼比对应的瞬态特性。针对实物平台中主导未知非线性较强、纯鲁棒项难以同时兼顾抖振抑制与跟踪精度的问题，提出神经网络预设阻尼比滑模反步控制方法。通过引入神经网络在线逼近主要未知非线性，建立权值更新律与权值误差系统，从而提高对主导未知非线性成分的补偿能力。

针对神经网络在线调整速度受限以及扰动作用下动态恢复品质不足的问题，提出二阶指令滤波抗扰预设阻尼比滑模反步控制方法。通过构造二阶指令滤波器及其状态空间表达，将滤波误差、名义拟合残差与外部扰动统一并入广义扰动项，建立相应误差状态方程并给出误差上界分析，从而提高系统对复合不确定性的动态拟合能力，并改善受扰后的恢复品质。

针对执行器幅值受限工况下输入饱和会破坏原有阻尼规律的问题，提出抗饱和预设阻尼比滑模反步控制扩展方法。通过在控制输入端引入抗饱和补偿动态，将饱和失配显式反馈到控制律中，抑制由输入受限引起的附加失配传播，从而减轻输入限幅对主导阻尼规律和跟踪性能的破坏。

最后，在四旋翼悬浮机平台上对所提三类方法进行了系统实验验证，完成了常值参考、时变参考、受扰工况和输入受限工况下的多组对比实验，并结合误差统计指标与实测数据反算阻尼比进行分析。结果表明：神经网络补偿后，预设阻尼比在强非线性实物平台上具有更好的可观察性；二阶指令滤波结构能够更快地对复合不确定性进行动态拟合并改善受扰恢复品质；抗饱和补偿能够减轻输入受限对主导阻尼规律的破坏，验证了所提方法的参数可解释性、工程可实现性和实物有效性。
\end{cabstract}

\begin{eabstract}
This thesis investigates predefined-damping-ratio control for uncertain nonlinear systems in strict-feedback form within a sliding-mode backstepping framework. Guided by a second-order Lyapunov method, the convergence of the output error is constrained by a second-order differential inequality, so that the dominant closed-loop dynamics can be directly shaped by controller parameters. In this way, transient specifications that are difficult to tune explicitly in conventional backstepping design are converted into interpretable damping-ratio-based parameter design rules.

First, for a class of uncertain strict-feedback nonlinear systems, a second-order Lyapunov stability criterion together with its parameter selection rules is established, and a predefined-damping-ratio sliding-mode backstepping controller is developed. By designing an integral sliding surface, virtual control laws, and the actual control input, a closed-loop error system with prescribed dominant damping parameters is obtained. It is shown that all closed-loop signals are uniformly ultimately bounded in the presence of unmodeled dynamics and external disturbances, while the output response approximately preserves the transient characteristics associated with the desired damping ratio.

Second, to address the strong dominant unknown nonlinearities on the real platform, for which a purely robust term cannot simultaneously suppress chattering and maintain tracking accuracy, a neural-network-based predefined-damping-ratio sliding-mode backstepping method is proposed. A neural network is introduced to approximate the dominant unknown nonlinearities online, and the associated weight-update law and weight-error system are derived. The approximation residual is then incorporated into the robust term, thereby improving the compensation capability for the dominant unknown nonlinear components.

Third, to overcome the limited adjustment speed of online neural-network adaptation and the insufficient recovery quality under disturbances, a second-order command-filtered anti-disturbance predefined-damping-ratio sliding-mode backstepping method is proposed. By constructing a second-order command filter and its state-space representation, the filter error, nominal fitting residual, and external disturbance are unified into a generalized disturbance term, from which the corresponding error-state equations and error-bound analysis are established. In this way, the dynamic fitting capability for composite uncertainties is improved and the disturbance-recovery performance is enhanced.

Finally, for actuator amplitude constraints that may destroy the originally designed damping behavior, an anti-saturation extension of the predefined-damping-ratio sliding-mode backstepping method is developed. An anti-saturation compensation dynamic is introduced so that the saturation-induced mismatch is explicitly fed back into the control law, thereby suppressing the additional performance degradation caused by input limitation.

Finally, systematic experimental validation is carried out on a quadrotor hover platform. Multiple comparative experiments are completed under constant references, time-varying references, disturbed conditions, and input-constrained conditions, and the measured data are further analyzed using error statistics and inverse damping-ratio estimation. The results show that neural-network compensation improves the observability of predefined damping ratios on a strongly nonlinear real platform, the command-filtered structure enables faster dynamic fitting of composite uncertainties and better disturbance-recovery performance, and the anti-saturation compensation alleviates the destruction of the dominant damping law caused by input constraints. Overall, the proposed methods exhibit good parameter interpretability, engineering implementability, and experimental effectiveness.
\end{eabstract}
